\section{Full Sheaf Construction}
\label{app:sheaf}

Let $G = (V, E)$ be a finite directed multigraph encoding a knowledge graph,
and let $\mathcal{O} = (T, R)$ be an Olog with type assignment
$\tau: V \to T$ and relation assignment $\rho: E \to R$. We construct a
cellular sheaf $\mathcal{F}$ on $G$ with $d$-dimensional stalks as follows.

\paragraph{Stalks.} For each vertex $v \in V$ we set
$\mathcal{F}(v) = \mathbb{R}^d$, and for each edge $e = (u \to v) \in E$ we
set $\mathcal{F}(e) = \mathbb{R}^d$. Where the KG admits 2-cells (triangles
in the underlying multigraph), we set $\mathcal{F}(\Delta) = \mathbb{R}^d$.

\paragraph{Restriction maps.} For each oriented edge $e = (u \to v)$,
choose linear maps $\mathcal{F}_{u \blacktriangleleft e}, \mathcal{F}_{v
\blacktriangleleft e} \in \mathbb{R}^{d \times d}$. We instantiate
$\mathcal{F}_{v \blacktriangleleft e}$ as the identity and
$\mathcal{F}_{u \blacktriangleleft e}$ as the relation embedding for
$\rho(e)$ (HDC binding under non-commutative random shifts; see
Appendix~\ref{app:impl}).

\paragraph{Cochain complex.} Set
\[
C^0 = \bigoplus_{v \in V} \mathcal{F}(v), \qquad
C^1 = \bigoplus_{e \in E} \mathcal{F}(e), \qquad
C^2 = \bigoplus_{\Delta} \mathcal{F}(\Delta).
\]
The coboundary $\delta_0: C^0 \to C^1$ is defined edge-wise by
$(\delta_0 x)_e = \mathcal{F}_{v \blacktriangleleft e} x_v -
\mathcal{F}_{u \blacktriangleleft e} x_u$. The coboundary
$\delta_1: C^1 \to C^2$ is the alternating sum of edge values around each
triangle $\Delta = (e_1, e_2, e_3)$.

\paragraph{Cohomology.}
\[
H^0(\mathcal{F}) = \ker \delta_0, \qquad
H^1(\mathcal{F}) = \ker \delta_1 / \mathrm{im}\,\delta_0.
\]
Standard arguments (see e.g.\ Hansen \& Gebhart, \cite{hansen2020sheaf})
verify $\delta_1 \delta_0 = 0$, so $H^1$ is well-defined.

\paragraph{Implementation and complexity.} We compute $\dim H^0$ as the
multiplicity of the zero eigenvalue of the sheaf Laplacian
$L = \delta_0^\top \delta_0 + \delta_1 \delta_1^\top$ within tolerance
$10^{-6}$ via dense symmetric eigensolve (\texttt{numpy.linalg.eigvalsh}),
and $\dim H^1 = \dim \ker \delta_1 - \mathrm{rank}\, \delta_0$ via dense
rank computation on the boundary matrices. With $n$ vertices and stalk
dimension $d$, this is $O((nd)^3)$ time and $O((nd)^2)$ space; tractable
for the type-graph view we use in our experiments ($n \leq 100$ Olog
objects per ontology), but not for the full entity-level KG. We note the
implementation uses sparse Krylov diffusion
(\texttt{scipy.sparse.linalg.expm\_multiply}) for query-time heat
propagation, but cohomology computation itself remains dense. A sparse
sheaf-cohomology algorithm via simultaneous Krylov-subspace methods on
$\delta_0$ and $\delta_1$ is a natural extension we leave to future work;
the cleaner alternative (computing the alternating sum
$\chi = \dim H^0 - \dim H^1 + \dim H^2$ via Euler characteristic in
$O(n + m)$) yields only the alternating sum, not individual Betti numbers,
when restriction maps are non-trivial.

\section{Proof of Theorem~\ref{thm:conflict_sensitivity}}
\label{app:conflict_proof}

We restate Theorem~\ref{thm:conflict_sensitivity} formally.

\begin{theorem}[Conflict Sensitivity, full statement]
\label{thm:conflict_sensitivity_full}
Let $\mathcal{F}$ be the cellular sheaf on $G = (V, E)$ defined in
Appendix~\ref{app:sheaf}. A \emph{conflict injection} of size $k$ is a set
$E^* = \{e^*_1, \ldots, e^*_k\}$ of additional edges with restriction maps
$\{(\mathcal{F}^*_{u_i \blacktriangleleft e^*_i}, \mathcal{F}^*_{v_i
\blacktriangleleft e^*_i})\}$ chosen so that for each $i$, no global section
$x \in \ker \delta_0$ of $\mathcal{F}$ satisfies
$\mathcal{F}^*_{v_i \blacktriangleleft e^*_i} x_{v_i} =
\mathcal{F}^*_{u_i \blacktriangleleft e^*_i} x_{u_i}$. Let $\mathcal{F}'$
denote the resulting sheaf on $G' = (V, E \cup E^*)$. Then
\begin{equation}
\dim H^1(\mathcal{F}') \;\geq\; \dim H^1(\mathcal{F}) + c,
\label{eq:conflict_sensitivity}
\end{equation}
where $c$ is the dimension of the subspace of $\bigoplus_{i=1}^{k}
\mathcal{F}(e^*_i)$ spanned by the conflict residuals
$r_i = \mathcal{F}^*_{v_i \blacktriangleleft e^*_i} x_{v_i} -
\mathcal{F}^*_{u_i \blacktriangleleft e^*_i} x_{u_i}$ evaluated on any
$\delta_0$-section, modulo $\mathrm{im}\,\delta'_0 + \mathrm{im}\,\delta'^\top_2$. Trivially, $c \leq k$.
\end{theorem}

\begin{proof}
The cochain spaces decompose as
\[
C^{0\prime} = C^0, \qquad
C^{1\prime} = C^1 \oplus \bigoplus_{i=1}^{k} \mathcal{F}(e^*_i), \qquad
C^{2\prime} = C^2 \oplus C^2_*
\]
where $C^2_*$ collects any new 2-cells closed by injected edges
(possibly empty).

Write $\delta'_0: C^{0\prime} \to C^{1\prime}$ in block form:
\[
\delta'_0 \;=\; \begin{pmatrix} \delta_0 \\ A \end{pmatrix},
\]
where $A: C^0 \to \bigoplus_i \mathcal{F}(e^*_i)$ is the restriction-map
contribution from the conflict edges, $A x = (r_1(x), \ldots, r_k(x))$.

The conflict hypothesis is precisely $A|_{\ker \delta_0} \neq 0$, i.e.\ there
is no $x \in \ker \delta_0$ with $A x = 0$ unless we shrink to a strict
subspace. Equivalently, $\ker \delta'_0 \subsetneq \ker \delta_0$, and
$\dim \ker \delta'_0 = \dim \ker \delta_0 - r_A$ where
$r_A = \mathrm{rank}\, A|_{\ker \delta_0}$.

By rank--nullity for $\delta'_0$:
\[
\mathrm{rank}\, \delta'_0 \;=\; \dim C^{0\prime} - \dim \ker \delta'_0
\;=\; \mathrm{rank}\, \delta_0 + r_A.
\]

For $H^1$:
\[
\dim H^1(\mathcal{F}') \;=\; \dim \ker \delta'_1 - \mathrm{rank}\, \delta'_0.
\]
We have $\dim \ker \delta'_1 \geq \dim \ker \delta_1 + k - \dim
\mathrm{im}\, \delta'^\top_2$ because each new edge contributes a new
generator to $C^{1\prime}$, which is killed by $\delta'_1$ unless it
participates in a new 2-cell relation in $C^2_*$.

Combining,
\begin{align*}
\dim H^1(\mathcal{F}') &\;\geq\; \dim \ker \delta_1 + k - \dim
\mathrm{im}\, \delta'^\top_2 - \mathrm{rank}\, \delta_0 - r_A \\
&\;=\; \dim H^1(\mathcal{F}) + (k - r_A - \dim \mathrm{im}\, \delta'^\top_2).
\end{align*}
Setting $c = k - r_A - \dim \mathrm{im}\, \delta'^\top_2$ gives
inequality~(\ref{eq:conflict_sensitivity}). The bound $c \leq k$ is
immediate.
\end{proof}

\paragraph{Empirical alignment.} Our experiment (Table~\ref{tab:conflicts})
gives $\Delta H^1 = +53$ for $k = 76$, hence $c/k \approx 0.70$. Roughly 70\%
of injected conflicts contribute linearly independent classes to $H^1$;
the remainder are absorbed by $\mathrm{im}\,\delta_0$ (i.e.\ correctable by
a global section adjustment) or by 2-cell boundaries.

\section{Implementation Details}
\label{app:impl}

The system has four components, each a separate module in the public
code release (Appendix~\ref{app:compute}).

\paragraph{HDC encoder.} Entities and relations are encoded as
4096-dimensional bipolar ($\pm 1$) hypervectors~\cite{kanerva2009hdc}.
Relation application uses non-commutative binding: each relation $r$ is
assigned a fixed random circular shift $\sigma_r$, so that $\sigma_r(h)
\neq \sigma_r^{-1}(h)$ and a directed edge $h \xrightarrow{r} t$ is
distinguishable from its reverse. The circular-shift (permutation)
binding follows the holographic reduced representation
family~\cite{plate1995hrr}.

\paragraph{Typed attention (B).} The (B) layer adds a precomputed mask
$M$ to the pre-softmax attention scores, with $M_{ij} = 0$ if
$\tau(j) \in \mathrm{Reach}_\mathcal{O}(\tau(i))$ and $M_{ij} = -\infty$
otherwise (in practice $-10^9$). The mask is built once per Olog from the
transitive closure of the morphism set.

\paragraph{Constrained decoder (D).} At each generation step the decoder
queries the proof object for the recommended step $(t, r, t')$,
constructs the admissible-label set $\mathcal{A}(t, r, t')$, and applies
a logit mask of $-\infty$ to every vocabulary entry outside it. The
decoder is stateless across steps apart from the proof object and
composes with (B) without interference.

\paragraph{Sheaf cohomology.} The cohomology module builds the boundary
operators $\delta_0, \delta_1$ from the restriction maps of
Appendix~\ref{app:sheaf} (identity on the head side, relation embedding
on the tail side) and computes $\dim H^1$ by dense rank on the boundary
matrices for the type-graph view, and by the $O(n+m)$ sparse
connected-components route for the entity-graph consistency score of
Table~\ref{tab:topology}.

\paragraph{Training configuration.} The attention ablation
(Table~\ref{tab:attention_ablation_b}, Figure~\ref{fig:typed_attention})
uses a two-head, 64-dimensional attention model trained 300 epochs with
Adam (learning rate 0.003) on next-type prediction over 23 types drawn
from four hand-built domain Ologs (business, academic, healthcare,
e-commerce). The HDC/Sheaf link-prediction model
(Table~\ref{tab:link_prediction}) uses 4096-dimensional embeddings
trained 50 epochs at batch size 512. All runs use seed 42.

\section{Hyperparameter Sweep}
\label{app:sweep}

We ran a 10-configuration sweep over HDC embedding dimension, learning
rate, and query-time diffusion time on FB15K-237, capped at 30 epochs
per configuration for rapid iteration (Table~\ref{tab:sweep}).

\begin{table}[t]
\centering
\caption{HDC/Sheaf hyperparameter sweep on FB15K-237, 30 epochs per
configuration, ranked by MRR. Diffusion time is a query-time heat-kernel
parameter and does not enter training; rows 3--6 (identical except for
diffusion time $\in \{0.5, 1.5, 3.0, 5.0\}$) produce identical ranking
metrics, confirming that retrieval accuracy is decoupled from the
diffusion parameter. The 30-epoch cap depresses absolute MRR relative to
the 50-epoch headline run of Table~\ref{tab:link_prediction} (0.346);
the sweep is for relative comparison only.}
\label{tab:sweep}
\begin{tabular}{clccc}
\toprule
Rank & HDC Dim / LR / Diff.\ Time & MRR & Hits@10 & Note \\
\midrule
1   & 1024 / 0.001 / 1.5      & \textbf{0.2284} & 0.3966 & best MRR \\
2   & 2048 / 0.001 / 1.5      & 0.2276 & \textbf{0.4024} & best Hits@10 \\
3--6 & 4096 / 0.001 / \{0.5--5.0\} & 0.2264 & 0.3989 & diffusion-invariant \\
7   & 4096 / 0.003 / 1.5      & 0.2259 & 0.3958 & \\
8   & 4096 / 0.0001 / 1.5     & 0.2257 & 0.3948 & lr too low \\
9   & 4096 / 0.010 / 1.5      & 0.2097 & 0.3742 & lr instability \\
10  & 8192 / 0.001 / 1.5      & 0.2096 & 0.3807 & underfit at 30 ep.\ \\
\bottomrule
\end{tabular}
\end{table}

Two findings. First, at a 30-epoch budget the lower dimensions
(1024--2048) converge faster and edge out 4096+, while 8192 underfits ---
larger hypervector spaces need more epochs to populate. Second, query
diffusion time has no effect on ranking metrics: it governs only the
heat-kernel spread used for path retrieval, not the embedding geometry.
Learning rate 0.001 is the stable choice; 0.01 induces optimization
instability and 0.0001 is too conservative.

\section{Cyclic Olog Stress Test}
\label{app:cyclic}

The cyclic stress test (Table~\ref{tab:b_prime_vs_d}, second row) takes
the 7-type e-commerce Olog and adds back-edges
(Delivery~$\to$~Customer, Item~$\to$~Customer) that close the otherwise
acyclic schema into a cycle. The effect is structural: with the cycle
present, every type can reach every other type, so the reachability
relation $\mathrm{Reach}_\mathcal{O}$ becomes the universal relation
(reachability rises from 34.7\% of type pairs to 100\%).

Because (B$'$) admits any next label whose target type is reachable, a
universal reachability relation makes the (B$'$) mask admit everything:
(B$'$) GOOD soundness falls from 65.0\% to 8.0\% and (B$'$) BAD from
3.9\% to 0.0\%, both within roughly one point of the unconstrained (A)
baseline (6.9\% / 0.0\%). The reachability mask is, at that point, doing
no work. The (D) constrained decoder is unaffected --- it masks to the
direct-edge admissible set $\mathcal{A}(t,r,t')$, which does not grow
when the Olog gains a cycle --- and remains at 100\% trajectory
soundness on both priors. This is the predicted topological limit:
reachability masking degrades to no-constraint as Olog connectivity
grows, whereas direct-edge masking is invariant to topology.

\section{Compute and Reproducibility}
\label{app:compute}

\paragraph{Hardware and runtime.} The HDC/Sheaf link-prediction model
trains on a single A100-40GB GPU (Modal), 50 epochs at batch size 512,
$\approx$30\,s/epoch ($\approx$25\,min end-to-end). The attention
ablation runs on CPU, $\approx$5\,min per model for 300 epochs. The loci
and cyclic-stress trajectory experiments run on CPU in seconds
($N=1000$ trajectories per cell). Type-graph sheaf cohomology is a
dense $O((nd)^3)$ computation but tractable at $n \leq 100$; the
entity-graph consistency score uses the $O(n+m)$ sparse route.

\paragraph{Code and seeds.} The implementation --- HDC encoder, typed
attention, constrained decoder, and the \texttt{compute\_cohomology}
routine --- is available at
\url{https://github.com/oasis-main/alignment_research} under
\texttt{ontological\_embeddings/}. All experiments use seed 42; the
attention ablation is reported single-seed and a multi-seed re-run is
noted as pending in \S\ref{sec:experiments}. Panel~(a) of
Figure~\ref{fig:typed_attention} is computed deterministically from the
Olog reachability closure; panel~(b) is read directly from the
attention-ablation results of Table~\ref{tab:attention_ablation_b}.

\section{(B) vs (D) --- Empirical Detail and Open Question}
\label{app:b_vs_d}

\paragraph{Empirical setup.} (B$'$), the sampling-time projection of (B),
admits a next label $\ell$ if its target type is reachable from the
current state under the Olog's morphism closure. We compare (A)
unconstrained, (B$'$) reachability-projected, (C) in-FFN compilation
(applicable only on functional sub-fragments), and (D) constrained
decoding on synthetic trajectories from the e-commerce Olog under both
GOOD and BAD label priors. Trajectory soundness is the fraction of
generated trajectories that decode to admissible $\delta_\mathcal{O}$-paths.

\paragraph{Result.} See Table~\ref{tab:b_prime_vs_d} in the main body
for the full numbers. Headlines: under acyclic Olog, (B$'$) GOOD reaches
65.0\% trajectory soundness vs (A) GOOD's 48.1\% and (D)'s 100\%;
under BAD prior (B$'$) sits at 3.9\% vs (A)'s 4.2\%, no meaningful
improvement. Under cyclic Olog (Delivery~$\to$~Customer added),
reachability becomes 100\% universal and (B$'$) collapses to 8.0\% GOOD
/ 0.0\% BAD, within a point of (A); (D) remains 100\%.

\paragraph{Open question (the curiosity case).} If (D) alone achieves
100\% trajectory soundness, what does (B) buy when (D) is also present?
Three hypotheses, none yet tested:
(i) \emph{representation prior:} (B) restricts information flow during
training, so the residual stream carries cleaner type information; (D)
alone leaves the residual unconstrained even if outputs are valid.
(ii) \emph{efficiency:} (B) reduces effective attention bandwidth during
forward pass, potentially saving compute on long sequences in exchange
for the per-layer mask cost.
(iii) \emph{robustness under decoder failure:} if (D) is removed (e.g.\
during research probing), a (B)-trained model degrades more gracefully
than a (B)-free one.
A four-cell ablation (A / B / D / B+D) trained from scratch on the same
data, with both downstream task accuracy and trajectory soundness
measured at convergence, would discriminate. We leave this to follow-up
work; the present paper reports (B) and (D) separately and does not
claim a Pareto win for (B)+(D) without it.

\paragraph{Implementation pointers.} (B) is the typed attention layer
described in \S\ref{sec:method:attention}; the production reference is
\texttt{OntologicalAttention} in
\texttt{ontological\_attention.py} (line 661,
\texttt{run\_slot\_ablation}). (D) is implemented in
\texttt{ConstrainedDecoder} of \texttt{proof\_guided\_generation.py}
(line 475). The decoder constructs the admissible-label set per
generation step from the proof object and applies a logit mask of
$-\infty$ to non-admissible vocabulary entries. Sheaf cohomology is
computed by the \texttt{compute\_cohomology} method of
\texttt{ontology\_sheaf.py} (line 219), which is now vendored into the
public repository.
